Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2024-12-12 manuscript_agu.tex
Grammar corrections

\begin{abstract}
Global storm-resolving models (GSRMs) are the upcoming global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). The high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but still utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100 km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using about 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and a station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate fog or near-surface clouds and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in the reanalyses, implying a ``too few, too bright'' cloud problem. Thermodynamic conditions are relatively well represented, but ICON is less stable than observations and MERRA-2 is too humid. SO cloud biases are a substantial issue in the GSRM, but it matches the observations better than the lower-resolution reanalyses.
\end{abstract}

@@

Increasing climate model resolution is one way of improving model accuracy of representation of the climate system \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution, is of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) is progressing to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) in research conditions today and operationally in the forthcoming years. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8 km \cite{satoh2019,stevens2019}. This is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1 August -- 10 September 2016) and winter (20 January -- 1 March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

@@

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net dowward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5 km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over high-latitude SO and the total cloud fraction underestimation on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48 hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

@@

The campaigns contained ceilometer observations captured by the Vaisala CL51 and CT25K, and the Lufft CHM 15k, described in detail below (Sections\ \ref{sec:cl51} and \ref{sec:chm15k}). A ceilometer is a low-power, near-infrared vertically pointing lidar principally designed to measure cloud base, but they also measure the full vertical structure of clouds as long as the laser signal is not attenuated by thick clouds, which can be used to infer additional information such as a cloud mask and cloud occurrence by height. We note that during the MICRE campaign, the ceilometers Vaisala CT25K and CL51 were installed at the Macquarie Island station concurrently, but in our analysis we only used the CT25K data obtained from the Atmospheric Radiation Measurement (ARM) data archive.

@@

TOA radiation quantities are taken from the CERES instruments onboard the Terra and Aqua satellites \cite{wielicki1996,loeb2018}. In our analysis we used the adjusted all-sky SW and LW upwelling fluxes at TOA from the synoptic TOA and surface fluxes and clouds 1-degree daily edition 4A product (CER\_SYN1deg-Day\_Terra-Aqua-MODIS\_Edition4A) \cite{doelling2013,doelling2016}.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/ann.pdf}
\caption{
Artificial neural network (ANN) for prediction of precipitation in lidar backscatter. \textbf{(a)} Diagram showing the TensorFlow structure of the ANN, \textbf{(b)} randomly selected example samples of near-surface backscatter in four categories (clear, fog, rain, and snow), as determined by coincident human-performed weather observations, \textbf{(c)} receiver operating characteristic diagram of the ANN, \textbf{(d)} examples of 10-day time series of human-observed (``HUM'') and predicted precipitation based on an ANN trained on all voyages (``ANN'') and all voyages except for the shown voyage (``ANN2'') during three randomly selected voyages with the available data. Here, by ``randomly selected'', we mean selected from the top of a permutation generated by a pseudo-random number generator to prevent authors' bias in the selection.
}
\label{fig:ann}
\end{figure}

@@

Precipitation can cause strong enough lidar backscattering to be recognized as clouds by the threshold-based cloud detection method used in the ALCF. This is undesirable if equivalent precipitation backscatter is not included in the simulated lidar profiles. It was not possible to include precipitation simulation in the ALCF due to the absence of required fields in the ICON model output and the reanalysis data (the liquid and ice precipitation mass mixing ratios). The required radiation calculations for precipitation are also currently not implemented in the ALCF, even though this is a planned future addition. In order to achieve a fair comparison of observations with model output, we exclude observed and simulated lidar profiles with precipitation either manually or using an automated method. It is relatively difficult to distinguish precipitation backscatter from cloud backscatter in lidar observations, especially when only one wavelength channel and no polarized channel are available. In models, the same can be accomplished relatively easily by excluding profiles exceeding a certain surface precipitation flux. In the observations, using precipitation flux measurements from rain gauges can be very unreliable on ships due to ship movement, turbulence caused by nearby ship structures, and sea spray. Our analysis of rain gauge data from the RV \emph{Tangaroa} showed large discrepancies between the rain gauge time series and human-performed synoptic observations, as well as large inconsistencies in the rain gauge time series. Human-performed observations of precipitation presence or absence are expected to be reliable but only cover a limited set of times. Therefore, it was desirable to implement a method of detecting precipitation from observed backscatter profiles alone.

@@

Fig.\ \ref{fig:cyclone-stability}a, b show a geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section\ \ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone, and it is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and the model agree within about 8\% in most areas, ICON is prevailingly more cyclonic by about 4\%. There are clear differences particularly in the highest occurrence rate regions such as around Cape Adare, which is by up to 20\% more cyclonic in ICON, and Weddel and Bellingshausen Sea, where ICON is less cyclonic by up to 10\%. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that the model is free-running.

Fig.\ \ref{fig:cyclone-stability}c, d show a geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section\ \ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (with respect to our SO partitioning; 50--65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are prevalent elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilising effect \cite{knight2024}. The ERA5 reanalysis is also substantially more stable than ICON across the whole region.

@@

When subsetted by conditions of weak and strong stability (Fig.\ \ref{fig:cloud-occurrence}f, g), as defined in Section\ \ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly nearly at the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1 km. It is worth mentioning that conditions of strong stability might be associated with the formation of advection fog, such as in situations of warm air advection from the north over colder sea surface thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak by up to 12\%, underestimated cloud occurrence above 1 km by up to 5\%, and underestimated cloud occurrence near the surface by about 11\%. In situations of weak stability, the bias in ICON is very pronounced, with a much stronger peak at about 500 m; ERA5 underestimates cloud occurrence below 1 km (especially near the surface); and MERRA-2 underestimates cloud occurrence even more strongly.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{img/cl_agg.pdf}
\caption{
Cloud occurrence by height calculated as the average of all voyages and stations and lifting condensation level (LCL) distribution. The LCL is derived from radiosonde profiles and equivalent model profiles, which were not available for all voyages and times. The total cloud fraction (CF), average shortwave (SW), and longwave (LW) and the relative frequency of occurrence (RFO) are shown. The bands are the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentile calculated from the set of all voyages and stations.
}
\label{fig:cloud-occurrence}
\end{figure}

@@

In the general case (Fig.\ \ref{fig:cloud-occurrence}a), ICON underestimates the outgoing SW radiation by 26 Wm$^{-2}$ and the reanalyses MERRA-2 and ERA5 are much closer, but overestimate it by 6 and 14 Wm$^{-2}$, respectively. While in ICON, this is in line with the underestimated total cloud fraction by 10\%, in the reanalyses this is the opposite result as expected from the underestimated total cloud fraction by about 20\%. The likely explanation is an overestimated cloud albedo, compensating for the lack of cloud area.

In contrast, LW radiation if affected by much smaller biases than SW radiation, which is expected due to the prevailing low-level clouds having similar temperature as the surface. In ICON, the outgoing LW radiation is overestimates by 8\%, which could be caused by understimated total cloud fraction exposing a larger area sea surface are to cooling to space, which is typically warmer than the atmospheric temperature at 0--2 km, where most of the clouds are located. In the MERRA-2 and ERA5 reanalyses, the LW biases are also slighly positive by 4 and 7 Wm$^{-2}$. This is again in line with the underestimated total cloud fraction by about 20\%. However, if the clouds are too thick, as expected from the SW results, this might also provide a compensating effect where too small cloud area is counteracted by greater thermal emissivity, thus reducing the ougoing LW radiation more relative to thinner clouds. For thin clouds, the outgoing TOA LW radiation originates both from the warmer surface (partly blocked by the clouds) and the clouds, whereas for thick clouds, the outgoing TOA LW radiation originates mostly from the colder-than-surface clouds.

In all of the subsets (Fig.\ \ref{fig:cloud-occurrence}b--g), the same type of bias is present as in the general case -- the outgoing SW radiation is underestimated in ICON and overestimated in MERRA-2 and ERA5, and the outgoing LW radiation is overestimated in all of the models. Even though the total cloud fraction is lower by 7\% over the high-latitude SO than the low-latitude SO, the outgoing SW radiation is much greater by 41 Wm$^{-2}$, implying a much greater cloud albedo over the high-latitude SO. ICON has the same contrast between these two regions in the total cloud fraction and SW radiation, but the outgoing SW radiation difference between the regions is much smaller (16 Wm$^{-2}$). The reanalyses do not show this type of contrast between the regions. The physical reason for this might be that the prevalence of fog or low-level clouds over the low-latitude SO and their relative lack over the high-latitude SO in obserations is not reproduced in the models (Fig.\ \ref{fig:cloud-occurrence}b--c).

@@

We also analyzed the daily cloud cover (total cloud fraction) distribution. This is a measure of cloudiness, irrespective of height, calculated over the course of a day (UTC). A cloud detected at any height means that the lidar profile was classified as cloudy; otherwise, it was classified as a clear sky. When all profiles in a day are taken together, the cloud cover for the day is defined as the fraction of cloudy profiles in the total number of profiles, expressed in oktas (multiples of 1/8). The same calculation is done the lidar observations as for the simulated lidar profiles. We use the term ``okta'' independently of its use in instantaneous synoptic observations, and here it simply means 1/8 (0.125\%) of the daily cloud cover.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=0.72\textwidth]{img/theta_hur.pdf}
\caption{
Virtual potential temperature (virt.\ pot.\ temp.; $\theta_v$) and relative humidity (RH) determined from radiosonde launches and co-located profiles in ICON, ERA5, and MERRA-2 in subsets as in Fig.\ \ref{fig:cloud-occurrence}. The solid lines are the average calculated from the averages of every individual voyage and station. The bands span the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentiles calculated from the distribution of the voyage and station averages. Shown is also the relative frequency of occurrence and the number of profiles in each subset.
}
\label{fig:potential-temperature}
\end{figure}

@@

The time period of ICON is relatively short, with only four full years of simulation available. Moreover, the simulation is free-running and ocean-coupled, which means that observations had to be temporally mapped to this time period (at the same time relative to the start of the year) for the comparison. For these reasons, one can expect the results to be slightly different due to reasons unrelated to model biases, such as different weather conditions, partially accounted for by the cyclone and stability subsetting, and the phase of climate oscillations such as the ENSO in the observations and the model. The interannual variability in cloud occurrence in ICON can be seen in Fig.\ \ref{fig:cloud-occurrence-panel}, where each year in ICON is represented by a separate line. The interannual variability tends to be substantially smaller than the biases, and thus is unlikely to have a stong impact on the main findings.

>>>>

\begin{abstract}
Global storm-resolving models (GSRMs) are the upcoming global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). The high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but still utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100 km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using about 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and a station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate fog or near-surface clouds, and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in the reanalyses, implying a ``too few, too bright'' cloud problem. Thermodynamic conditions are relatively well represented, but ICON is less stable than observations and MERRA-2 is too humid. SO cloud biases are a substantial issue in the GSRM, but it matches the observations better than the lower-resolution reanalyses.
\end{abstract}

@@

Increasing climate model resolution is one way of improving model accuracy of representation of the climate system \cite{mauritsen2022}. It has been practiced since the advent of climate modeling as more computational power, memory, and storage capacity become available. It is, however, often not as easy as changing the grid size because of the complex interplay between model dynamics and physics, which necessitates adjusting and tuning all components together. Increasing resolution is, of course, limited by the available computational power and a trade-off with increasing parameterization complexity, which is another way of improving model accuracy. Current computational availability and acceleration from general-purpose computing on graphics processing units (GPUs) is progressing to enable km-scale (also called k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general circulation models (AOGCMs) in research conditions today and operationally in the forthcoming years. Therefore, it represents a natural advance in climate modeling. Global storm-resolving models (GSRMs) are emerging as a new front in the development of high-resolution global climate models, with horizontal grid resolutions of about 2--8 km \cite{satoh2019,stevens2019}. This is enough to resolve mesoscale convective storms, but smaller-scale convective plumes and cloud structure remain unresolved. At an approximately 5-km scale, non-hydrostatic processes also become important \cite{weisman1997}, and for this reason such models are generally non-hydrostatic. The terms global cloud-resolving models or global convection-permitting/-resolving models are also sometimes used interchangeably with GSRMs but imply that clouds or convection are resolved explicitly, which is not entirely true for GSRMs, as this would require an even higher horizontal resolution \cite{satoh2019}. Representative of these efforts is the DYnamics of the Atmospheric general circulation Modeled On Non-hydrostatic Domains (DYAMOND) project \cite{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs over two 40-day time periods in summer (1 August -- 10 September 2016) and winter (20 January -- 1 March 2020). A new one-year GSRM intercomparison is currently proposed by \citeA{takasuka2024}, with the hope of also evaluating the seasonal cycle and large-scale circulation. An alternative to using a computationally costly GSRM is to train an artificial neural network on GSRM output and use it for subgrid-scale clouds, as done with the GSRM ICON by \citeA{grundner2022} and \citeA{grundner2023}.

@@

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight analyzed CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small cloud fraction and too large cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO) Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5 km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over high-latitude SO and the total cloud fraction underestimation on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48 hours of km-scale ICON limited-area model NWP simulations over a SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

@@

The campaigns contained ceilometer observations captured by the Vaisala CL51 and CT25K, and the Lufft CHM 15k, described in detail below (Sections\ \ref{sec:cl51} and \ref{sec:chm15k}). A ceilometer is a low-power, near-infrared, vertically pointing lidar principally designed to measure cloud base, but they also measure the full vertical structure of clouds as long as the laser signal is not attenuated by thick clouds, which can be used to infer additional information such as a cloud mask and cloud occurrence by height. We note that during the MICRE campaign, the ceilometers Vaisala CT25K and CL51 were installed at the Macquarie Island station concurrently, but in our analysis we only used the CT25K data obtained from the Atmospheric Radiation Measurement (ARM) data archive.

@@

TOA radiation quantities are taken from the CERES instruments onboard the Terra and Aqua satellites \cite{wielicki1996,loeb2018}. In our analysis, we used the adjusted all-sky SW and LW upwelling fluxes at TOA from the synoptic TOA and surface fluxes and clouds 1-degree daily edition 4A product (CER\_SYN1deg-Day\_Terra-Aqua-MODIS\_Edition4A) \cite{doelling2013,doelling2016}.

@@

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/ann.pdf}
\caption{
Artificial neural network (ANN) for prediction of precipitation in lidar backscatter. \textbf{(a)} Diagram showing the TensorFlow structure of the ANN, \textbf{(b)} randomly selected example samples of near-surface backscatter in four categories (clear, fog, rain, and snow), as determined by coincident human-performed weather observations, \textbf{(c)} receiver operating characteristic diagram of the ANN, \textbf{(d)} examples of 10-day time series of human-observed (``HUM'') and predicted precipitation based on an ANN trained on all voyages (``ANN'') and all voyages except for the shown voyage (``ANN2'') during three randomly selected voyages with the available data. Here, by ``randomly selected,'' we mean selected from the top of a permutation generated by a pseudo-random number generator to prevent authors' bias in the selection.
}
\label{fig:ann}
\end{figure}

@@

Precipitation can cause strong enough lidar backscattering to be recognized as clouds by the threshold-based cloud detection method used in the ALCF. This is undesirable if equivalent precipitation backscatter is not included in the simulated lidar profiles. It was not possible to include precipitation simulation in the ALCF due to the absence of required fields in the ICON model output and the reanalysis data (the liquid and ice precipitation mass mixing ratios). The required radiation calculations for precipitation are also currently not implemented in the ALCF, even though this is a planned future addition. In order to achieve a fair comparison of observations with model output, we exclude observed and simulated lidar profiles with precipitation, either manually or using an automated method. It is relatively difficult to distinguish precipitation backscatter from cloud backscatter in lidar observations, especially when only one wavelength channel and no polarized channel are available. In models, the same can be accomplished relatively easily by excluding profiles exceeding a certain surface precipitation flux. In the observations, using precipitation flux measurements from rain gauges can be very unreliable on ships due to ship movement, turbulence caused by nearby ship structures, and sea spray. Our analysis of rain gauge data from the RV \emph{Tangaroa} showed large discrepancies between the rain gauge time series and human-performed synoptic observations, as well as large inconsistencies in the rain gauge time series. Human-performed observations of precipitation presence or absence are expected to be reliable but only cover a limited set of times. Therefore, it was desirable to implement a method of detecting precipitation from observed backscatter profiles alone.

@@

Fig.\ \ref{fig:cyclone-stability}a, b show a geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section\ \ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone, and it is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and the model agree within about 8\% in most areas, ICON is prevailingly more cyclonic by about 4\%. There are clear differences, particularly in the highest occurrence rate regions, such as around Cape Adare, which is by up to 20\% more cyclonic in ICON, and the Weddell and Bellingshausen Seas, where ICON is less cyclonic by up to 10\%. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that the model is free-running.

Fig.\ \ref{fig:cyclone-stability}c, d show a geographical distribution of the conditions of weak and strong stability as determined by the LTS (Section\ \ref{sec:cyclone-stability}). Conditions of weak stability are prevalent in the mid-to-high SO (with respect to our SO partitioning; 50--65°S), which might be explained by the relatively cold near-surface air overlying the relatively warm sea surface. Conditions of strong stability are prevalent elsewhere over the SO. The distribution is also less zonally symmetric than the cyclonic activity. In the high-latitude SO, the presence of sea ice might have a substantial stabilizing effect \cite{knight2024}. The ERA5 reanalysis is also substantially more stable than ICON across the whole region.

@@

When subsetted by conditions of weak and strong stability (Fig.\ \ref{fig:cloud-occurrence}f, g), as defined in Section\ \ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly nearly at the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1 km. It is worth mentioning that conditions of strong stability might be associated with the formation of advection fog, such as in situations of warm air advection from the north over a colder sea surface, thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak by up to 12\%, underestimated cloud occurrence above 1 km by up to 5\%, and underestimated cloud occurrence near the surface by about 11\%. In situations of weak stability, the bias in ICON is very pronounced, with a much stronger peak at about 500 m; ERA5 underestimates cloud occurrence below 1 km (especially near the surface); and MERRA-2 underestimates cloud occurrence even more strongly.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{img/cl_agg.pdf}
\caption{
Cloud occurrence by height calculated as the average of all voyages and stations and lifting condensation level (LCL) distribution. The LCL is derived from radiosonde profiles and equivalent model profiles, which were not available for all voyages and times. The total cloud fraction (CF), average shortwave (SW), and longwave (LW) and the relative frequency of occurrence (RFO) are shown. The bands are the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentile, calculated from the set of all voyages and stations.
}
\label{fig:cloud-occurrence}
\end{figure}

@@

In the general case (Fig.\ \ref{fig:cloud-occurrence}a), ICON underestimates the outgoing SW radiation by 26 Wm$^{-2}$ and the reanalyses MERRA-2 and ERA5 are much closer but overestimate it by 6 and 14 Wm$^{-2}$, respectively. While in ICON, this is in line with the underestimated total cloud fraction by 10\%, in the reanalyses this is the opposite result as expected from the underestimated total cloud fraction by about 20\%. The likely explanation is an overestimated cloud albedo, compensating for the lack of cloud area.

In contrast, LW radiation is affected by much smaller biases than SW radiation, which is expected due to the prevailing low-level clouds having similar temperature as the surface. In ICON, the outgoing LW radiation is overestimated by 8\%, which could be caused by an underestimated total cloud fraction exposing a larger sea surface area to cooling to space, which is typically warmer than the atmospheric temperature at 0--2 km, where most of the clouds are located. In the MERRA-2 and ERA5 reanalyses, the LW biases are also slightly positive by 4 and 7 Wm$^{-2}$. This is again in line with the underestimated total cloud fraction by about 20\%. However, if the clouds are too thick, as expected from the SW results, this might also provide a compensating effect, in which too small a cloud area is counteracted by greater thermal emissivity, thus reducing the outgoing LW radiation more relative to thinner clouds. For thin clouds, the outgoing TOA LW radiation originates both from the warmer surface (partly blocked by the clouds) and the clouds, whereas for thick clouds, the outgoing TOA LW radiation originates mostly from the colder-than-surface clouds.

In all the subsets (Fig.\ \ref{fig:cloud-occurrence}b--g), the same type of bias is present as in the general case -- the outgoing SW radiation is underestimated in ICON and overestimated in MERRA-2 and ERA5, and the outgoing LW radiation is overestimated in all the models. Even though the total cloud fraction is lower by 7\% over the high-latitude SO than the low-latitude SO, the outgoing SW radiation is much greater by 41 Wm$^{-2}$, implying a much greater cloud albedo over the high-latitude SO. ICON has the same contrast between these two regions in the total cloud fraction and SW radiation, but the outgoing SW radiation difference between the regions is much smaller (16 Wm$^{-2}$). The reanalyses do not show this type of contrast between the regions. The physical reason for this might be that the prevalence of fog or low-level clouds over the low-latitude SO and their relative lack over the high-latitude SO in observations is not reproduced in the models (Fig.\ \ref{fig:cloud-occurrence}b--c).

@@

We also analyzed the daily cloud cover (total cloud fraction) distribution. This is a measure of cloudiness, irrespective of height, calculated over the course of a day (UTC). A cloud detected at any height means that the lidar profile was classified as cloudy; otherwise, it was classified as a clear sky. When all profiles in a day are taken together, the cloud cover for the day is defined as the fraction of cloudy profiles in the total number of profiles, expressed in oktas (multiples of 1/8). The same calculation is done for the lidar observations as for the simulated lidar profiles. We use the term ``okta'' independently of its use in instantaneous synoptic observations, and here it simply means 1/8 (0.125\%) of the daily cloud cover.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=0.72\textwidth]{img/theta_hur.pdf}
\caption{
Virtual potential temperature (virt.\ pot.\ temp.; $\theta_v$) and relative humidity (RH) determined from radiosonde launches and co-located profiles in ICON, ERA5, and MERRA-2 in subsets as in Fig.\ \ref{fig:cloud-occurrence}. The solid lines are the average calculated from the averages of every individual voyage and station. The bands span the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentiles, calculated from the distribution of the voyage and station averages. Shown is also the relative frequency of occurrence and the number of profiles in each subset.
}
\label{fig:potential-temperature}
\end{figure}

@@

The time period of ICON is relatively short, with only four full years of simulation available. Moreover, the simulation is free-running and ocean-coupled, which means that observations had to be temporally mapped to this time period (at the same time relative to the start of the year) for the comparison. For these reasons, one can expect the results to be slightly different due to reasons unrelated to model biases, such as different weather conditions, partially accounted for by the cyclone and stability subsetting, and the phase of climate oscillations such as the ENSO in the observations and the model. The interannual variability in cloud occurrence in ICON can be seen in Fig.\ \ref{fig:cloud-occurrence-panel}, where each year in ICON is represented by a separate line. The interannual variability tends to be substantially smaller than the biases, and thus is unlikely to have a strong impact on the main findings.
